---
title: "Equilibrium Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Equilibrium Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 6,
  fig.height = 5,
  fig.align  = "center",
  out.width  = "90%"
)
```

## Introduction

An **equilibrium point** (also called a fixed point or steady state) of
an autonomous ODE system is a state $(x^*, y^*)$ where all derivatives
vanish simultaneously:

$$f(x^*, y^*) = 0, \qquad g(x^*, y^*) = 0.$$

Knowing the location and **stability type** of each equilibrium is
central to understanding the qualitative behavior of a system — it
tells us where trajectories end up, how they approach or depart from
rest states, and how the system responds to perturbations.

This vignette covers:

1. Finding equilibria numerically with `find_equilibrium()`
2. The Jacobian matrix and linearization
3. Classifying equilibria with `classify_equilibrium()`
4. The trace-determinant plane
5. Drawing stable and unstable manifolds with `gg_manifolds()`
6. A complete worked example: the nonlinear competition model
7. Bifurcation preview: how equilibrium type changes with parameters

```{r load-packages, message = FALSE}
library(ggphasr)
library(ggplot2)
```

---

## 1  Finding equilibria numerically

For simple systems, equilibria can often be found by hand.  For
nonlinear systems, numerical root-finding is more reliable.
`find_equilibrium()` uses Newton–Raphson iteration (via
`rootSolve::multiroot()`) to locate equilibria from one or more
starting points.

### 1.1  Single starting point

```{r find-single}
lv_params <- c(alpha = 1, beta = 0.5, delta = 0.5, gamma = 1)

# Known interior equilibrium near (2, 2)
eq_interior <- find_equilibrium(
  ode_lotka_volterra,
  y0         = c(1.5, 1.5),
  parameters = lv_params
)
eq_interior
```

### 1.2  Grid search for multiple equilibria

When a system has several equilibria and their locations are not known
in advance, use `y0 = NULL` to trigger an automatic grid search.
`find_equilibrium()` places starting points on a regular grid over
`xlim` × `ylim`, runs Newton–Raphson from each, and deduplicates the
results:

```{r find-grid}
# example_11: dx/dt = x(3-x-2y), dy/dt = y(2-x-y)
# Four equilibria: (0,0), (3,0), (0,2), (1,1)
eq_list <- find_equilibrium(
  ode_example_11,
  y0     = NULL,
  xlim   = c(0, 4),
  ylim   = c(0, 3),
  n_grid = 12L
)

# Print each equilibrium rounded to 4 decimal places
lapply(eq_list, round, digits = 4)
```

> **Technical aside:** Newton–Raphson requires an initial guess and may
> converge to a non-equilibrium point or fail to converge if the
> starting point is far from any root.  `find_equilibrium()` guards
> against this by verifying that the residual $\|f(x^*)\|_\infty$ is
> below a tolerance threshold before accepting a root.  Results are
> sorted by x-coordinate and deduplicated using a Euclidean distance
> threshold (`dedup_tol = 1e-4` by default).

---

## 2  The Jacobian matrix and linearization

The stability of an equilibrium $(x^*, y^*)$ is determined by the
behavior of nearby solutions.  Near an equilibrium, the nonlinear
system

$$\frac{d\mathbf{y}}{dt} = \mathbf{F}(\mathbf{y})$$

can be approximated by its **linearization**:

$$\frac{d\boldsymbol{\xi}}{dt} \approx J\,\boldsymbol{\xi},$$

where $\boldsymbol{\xi} = \mathbf{y} - \mathbf{y}^*$ is the
displacement from equilibrium and $J$ is the **Jacobian matrix**:

$$J = \begin{pmatrix}
\partial f/\partial x & \partial f/\partial y \\
\partial g/\partial x & \partial g/\partial y
\end{pmatrix}\bigg|_{(x^*, y^*)}.$$

The eigenvalues $\lambda_1, \lambda_2$ of $J$ determine the local
behavior:

- If both $\text{Re}(\lambda_i) < 0$: the equilibrium is **stable**
  (trajectories approach it).
- If any $\text{Re}(\lambda_i) > 0$: the equilibrium is **unstable**
  (trajectories depart from it).
- If $\lambda_1 < 0 < \lambda_2$ (one of each sign): the equilibrium
  is a **saddle**.

`classify_equilibrium()` computes the Jacobian numerically via forward
finite differences and classifies the equilibrium using the
**trace-determinant** criterion (see Section 4):

```{r jacobian-example}
# Jacobian at the Lotka-Volterra interior equilibrium (2, 2)
# Analytic Jacobian: [[0, -1], [0.5, 0]] -> tr = 0, det = 0.5
# -> Center (purely imaginary eigenvalues)

cl_interior <- classify_equilibrium(
  ode_lotka_volterra,
  equilibrium = eq_interior[[1L]],
  parameters  = lv_params
)

cl_interior[, c("x", "y", "classification", "tr", "det",
                 "jacobian_11", "jacobian_12",
                 "jacobian_21", "jacobian_22")]
```

```{r jacobian-eigenvalues}
cl_interior[, c("lambda_1_re", "lambda_1_im",
                 "lambda_2_re", "lambda_2_im")]
```

The eigenvalues are $\lambda = \pm i\sqrt{0.5} \approx \pm 0.707\,i$
— purely imaginary, confirming the center classification.

---

## 3  Classification table for a complete system

Combining `find_equilibrium()` and `classify_equilibrium()` via
`lapply()` produces a tidy summary table for all equilibria at once:

```{r classify-all}
# All four equilibria of example_11
eq_table_11 <- do.call(rbind, lapply(eq_list, function(eq) {
  classify_equilibrium(ode_example_11, equilibrium = eq)
}))

eq_table_11[, c("x", "y", "classification", "tr", "det")]
```

This table immediately reveals the qualitative structure: the origin
and the coexistence point $(1, 1)$ are saddles, while $(3, 0)$ and
$(0, 2)$ are the stable single-species equilibria.

---

## 4  The trace-determinant plane

For a 2D system, the Jacobian has two eigenvalues whose character is
completely determined by the **trace** $\tau = \text{tr}(J)$ and
**determinant** $\Delta = \det(J)$:

$$\lambda_{1,2} = \frac{\tau \pm \sqrt{\tau^2 - 4\Delta}}{2}.$$

The classification rules are:

| Region | Condition | Type |
|---|---|---|
| $\Delta < 0$ | Any $\tau$ | Saddle |
| $\Delta > 0$, $\tau^2 - 4\Delta > 0$, $\tau < 0$ | — | Stable node |
| $\Delta > 0$, $\tau^2 - 4\Delta > 0$, $\tau > 0$ | — | Unstable node |
| $\Delta > 0$, $\tau^2 - 4\Delta < 0$, $\tau < 0$ | — | Stable spiral |
| $\Delta > 0$, $\tau^2 - 4\Delta < 0$, $\tau > 0$ | — | Unstable spiral |
| $\Delta > 0$, $\tau = 0$ | — | Center |

The parabola $\tau^2 = 4\Delta$ separates nodes from spirals.

We can visualize this by tracking how the equilibrium type of the
**simple pendulum's stable equilibrium** at the origin moves through
the trace-determinant plane as the damping coefficient $b$ increases:

```{r td-plane, fig.height = 4.5}
b_vals   <- seq(0, 3, by = 0.15)

td_table <- do.call(rbind, lapply(b_vals, function(b) {
  result    <- classify_equilibrium(
    ode_simple_pendulum,
    equilibrium = c(0, 0),
    parameters  = c(gL = 1, b = b)
  )
  result$b  <- b
  result
}))

# Parabola tau^2 = 4*delta (node/spiral boundary)
tau_seq    <- seq(-3.2, 0.2, length.out = 200)
parabola   <- data.frame(tau = tau_seq, delta = tau_seq^2 / 4)

ggplot(td_table, aes(x = tr, y = det)) +
  # Parabola separating nodes from spirals
  geom_line(data = parabola, aes(x = tau, y = delta),
            color = "grey60", linetype = "dashed", linewidth = 0.7) +
  # Reference lines
  geom_hline(yintercept = 0, color = "grey70", linewidth = 0.5) +
  geom_vline(xintercept = 0, color = "grey70", linewidth = 0.5) +
  # Path through the plane as b increases
  geom_path(color = "grey50", linewidth = 0.6) +
  geom_point(aes(color = classification), size = 3) +
  scale_color_brewer(palette = "Set1", name = "Type") +
  # Annotate key regions
  annotate("text", x = -2.5, y = 2.5,
           label = "Stable\nspiral",   size = 3, color = "grey40") +
  annotate("text", x = -2.5, y = 0.3,
           label = "Stable\nnode",     size = 3, color = "grey40") +
  annotate("text", x =  0.5, y = 2.5,
           label = "Unstable\nspiral", size = 3, color = "grey40") +
  annotate("text", x = -1.0, y = -0.5,
           label = "Saddle",           size = 3, color = "grey40") +
  annotate("text", x =  0.1, y = 1.05,
           label = "tau^2 == 4*Delta", parse = TRUE,
           size  = 3, color = "grey50", hjust = 0) +
  labs(
    x     = "Trace (tau)",
    y     = "Determinant (Delta)",
    title = "Trace-determinant plane: pendulum origin as b increases"
  ) +
  coord_cartesian(xlim = c(-3.5, 1), ylim = c(-0.8, 3)) +
  theme_phase_plane()
```

As $b$ increases from 0, the equilibrium moves from a **center**
($\tau = 0$, $\Delta > 0$) through **stable spirals** ($\tau < 0$,
complex eigenvalues) and crosses the parabola $\tau^2 = 4\Delta$ to
become a **stable node** (real eigenvalues) — the overdamped regime.

---

## 5  Stable and unstable manifolds

At a **saddle point**, the Jacobian has one positive and one negative
eigenvalue.  The corresponding eigenvectors define two special
trajectories:

- **Stable manifold** $W^s$: the set of initial conditions that
  converge to the saddle as $t \to +\infty$.  Locally aligned with
  the eigenvector of the negative eigenvalue.
- **Unstable manifold** $W^u$: the set of initial conditions that
  converge to the saddle as $t \to -\infty$ (equivalently, depart
  from it as $t \to +\infty$).  Locally aligned with the eigenvector
  of the positive eigenvalue.

`gg_manifolds()` computes both manifolds by:

1. Computing the Jacobian eigenvectors at the saddle.
2. Placing seed points at distance `epsilon` from the saddle along
   each eigenvector direction.
3. Integrating **forward** from the unstable seeds and **backward**
   from the stable seeds.

### 5.1  Linear saddle: example_08

For the linear system $dx/dt = x$, $dy/dt = -y$, the stable manifold
is exactly the $y$-axis and the unstable manifold is exactly the
$x$-axis — a useful verification case:

```{r manifolds-linear}
gg_flow_field(ode_example_08,
              xlim  = c(-3, 3),
              ylim  = c(-3, 3),
              title = "Linear saddle: exact manifolds") +
  gg_manifolds(ode_example_08,
               equilibrium = c(0, 0))
```

### 5.2  Nonlinear saddle: example_11 competition model

For the nonlinear competition model, the saddle at $(1, 1)$ has curved
manifolds.  The **stable manifold is the separatrix** — the curve that
divides the phase plane into the basin of attraction of $(3, 0)$
(species 1 wins) and the basin of attraction of $(0, 2)$ (species 2
wins):

```{r manifolds-competition}
# First classify the saddle to reuse the Jacobian
eq_saddle <- find_equilibrium(ode_example_11, y0 = c(0.8, 0.8))
cl_saddle  <- classify_equilibrium(ode_example_11,
                                    equilibrium = eq_saddle[[1L]])
cl_saddle[, c("x", "y", "classification")]

gg_flow_field(ode_example_11,
              xlim  = c(0, 4),
              ylim  = c(0, 3),
              title = "Competition model: saddle manifolds") +
  gg_nullclines(ode_example_11,
                xlim            = c(0, 4),
                ylim            = c(0, 3),
                legend_position = "bottom") +
  gg_manifolds(ode_example_11,
               equilibrium   = eq_saddle[[1L]],
               eq_classified = cl_saddle,
               t_manifold    = 5) +
  gg_trajectory(ode_example_11,
                y0   = list(c(0.2, 0.2), c(3.5, 0.2),
                             c(0.2, 2.8), c(2.0, 2.0)),
                xlim = c(0, 4), ylim = c(0, 3),
                t_end = 8, color = "grey50",
                add_start_point = FALSE, add_arrows = FALSE)
```

Trajectories starting on opposite sides of the stable manifold (blue
dashed) converge to different outcomes — a concrete illustration of
why the separatrix is so important for understanding long-run behavior.

---

## 6  Complete worked example: genetic toggle switch

The Gardner et al. (2000) genetic toggle switch models two mutually
repressing proteins $u$ and $v$:

$$\frac{du}{dt} = \frac{\alpha_1}{1 + v^\beta} - u, \qquad
\frac{dv}{dt} = \frac{\alpha_2}{1 + u^\gamma} - v.$$

For sufficiently large $\alpha_1$, $\alpha_2$ and cooperativity
coefficients $\beta$, $\gamma > 1$, the system is **bistable**: two
stable equilibria (one protein dominant) separated by an unstable
saddle.  This bistability is the mechanism of synthetic biological
memory.

```{r toggle-params}
toggle_params <- c(alpha1 = 3, alpha2 = 3, beta = 2, gamma = 2)
```

### 6.1  Phase plane and equilibria

```{r toggle-phase}
result_toggle <- gg_phase_plane(
  ode_toggle,
  xlim            = c(0, 4),
  ylim            = c(0, 4),
  parameters      = toggle_params,
  t_end           = 10,
  n_ic            = 5,
  title           = "Genetic toggle switch",
  legend_position = "bottom"
)
result_toggle$plot +
  labs(x = "Protein u", y = "Protein v")
```

```{r toggle-eq}
result_toggle$equilibria[, c("x", "y", "classification", "tr", "det")]
```

The toggle switch has three equilibria: two stable nodes (one protein
dominant) and one saddle (unstable coexistence).

### 6.2  Separatrix via manifolds

```{r toggle-manifolds}
# Extract the saddle
saddle_toggle <- result_toggle$equilibria[
  result_toggle$equilibria$classification == "Saddle", ]
saddle_coords <- c(saddle_toggle$x[[1L]], saddle_toggle$y[[1L]])

# Reclassify to get the Jacobian
cl_toggle <- classify_equilibrium(ode_toggle,
                                   equilibrium = saddle_coords,
                                   parameters  = toggle_params)

result_toggle$plot +
  gg_manifolds(ode_toggle,
               equilibrium   = saddle_coords,
               parameters    = toggle_params,
               eq_classified = cl_toggle,
               t_manifold    = 6) +
  labs(x     = "Protein u",
       y     = "Protein v",
       title = "Toggle switch: separatrix")
```

The stable manifold (blue dashed curve) is the separatrix between the
two basins of attraction.  A perturbation that pushes the system across
this boundary causes an irreversible switch from one stable state to
the other — this is precisely the "memory" property of the toggle
switch.

---

## 7  Bifurcation preview: how equilibria change with parameters

A **bifurcation** occurs when a qualitative change in the phase
portrait happens as a parameter crosses a critical value.  `ggphasr`
can illustrate bifurcations by repeating the classification workflow
across a range of parameter values.

### 7.1  Hopf bifurcation in the Van der Pol oscillator

The Van der Pol origin is an unstable spiral for $\mu > 0$ and a
center for $\mu = 0$.  As $\mu$ increases, the trace of the Jacobian
increases, moving the origin further into the unstable region.  We can
track this:

```{r hopf-table}
mu_vals <- c(0, 0.5, 1, 2, 3)

hopf_table <- do.call(rbind, lapply(mu_vals, function(mu) {
  result    <- classify_equilibrium(
    ode_van_der_pol,
    equilibrium = c(0, 0),
    parameters  = c(mu = mu)
  )
  result$mu <- mu
  result
}))

hopf_table[, c("mu", "classification", "tr", "det",
                "lambda_1_re", "lambda_1_im")]
```

At $\mu = 0$ the origin is a center ($\tau = 0$).  For $\mu > 0$ the
trace becomes positive and the origin is an unstable spiral — a Hopf
bifurcation occurs at $\mu = 0$.  For $\mu > 0$ a stable limit cycle
exists (as seen in the previous vignette).

### 7.2  Saddle-node bifurcation preview

Consider a user-defined 1D system with a bifurcation parameter $r$:

$$\frac{dy}{dt} = r + y^2.$$

For $r < 0$ there are two equilibria at $y^* = \pm\sqrt{-r}$.  At
$r = 0$ they collide in a **saddle-node bifurcation** and disappear
for $r > 0$.

```{r sn-bifurcation}
saddle_node <- function(y, parameters = NULL) {
  r <- parameters[["r"]]
  r + y^2
}

# r = -1: two equilibria
eq_sn_neg <- find_equilibrium(saddle_node,
                               y0 = NULL, system = "one.dim",
                               ylim = c(-3, 3),
                               parameters = c(r = -1))
cat("r = -1: equilibria at y =",
    round(sapply(eq_sn_neg, `[[`, 1L), 4), "\n")

# r = 0: one equilibrium (the bifurcation point)
eq_sn_zero <- find_equilibrium(saddle_node,
                                y0 = NULL, system = "one.dim",
                                ylim = c(-1, 1),
                                parameters = c(r = 0))
cat("r = 0:  equilibria at y =",
    round(sapply(eq_sn_zero, `[[`, 1L), 4), "\n")

# r = 1: no equilibria in this range
eq_sn_pos <- find_equilibrium(saddle_node,
                               y0 = NULL, system = "one.dim",
                               ylim = c(-3, 3),
                               parameters = c(r = 1))
cat("r = 1:  equilibria found:", length(eq_sn_pos), "\n")
```

```{r sn-portraits, fig.width = 9, fig.height = 3.5}
make_sn_plot <- function(r) {
  gg_flow_field(
    saddle_node,
    xlim       = c(0, 3),
    ylim       = c(-2.5, 2.5),
    system     = "one.dim",
    parameters = c(r = r),
    title      = paste0("r = ", r)
  ) +
    gg_phase_portrait(
      saddle_node,
      ylim       = c(-2.5, 2.5),
      xlim       = c(0, 3),
      parameters = c(r = r)
    )
}

if (requireNamespace("patchwork", quietly = TRUE)) {
  library(patchwork)
  make_sn_plot(-1) | make_sn_plot(0) | make_sn_plot(1)
} else {
  make_sn_plot(-1)
}
```

The three panels show the phase line before the bifurcation (two
equilibria), at the bifurcation (one semi-stable equilibrium), and
after the bifurcation (no equilibria — all solutions escape to
$+\infty$).

---

## Summary

This vignette covered the complete equilibrium analysis workflow in
`ggphasr`:

| Task | Function |
|---|---|
| Find equilibria (single guess) | `find_equilibrium(deriv, y0 = c(...))` |
| Find equilibria (grid search) | `find_equilibrium(deriv, y0 = NULL, xlim, ylim)` |
| Classify one equilibrium | `classify_equilibrium(deriv, equilibrium)` |
| Classify all equilibria | `do.call(rbind, lapply(eq_list, classify_equilibrium, ...))` |
| Draw manifolds | `gg_manifolds(deriv, equilibrium, eq_classified)` |
| All-in-one analysis | `gg_phase_plane(deriv, xlim, ylim)` |

The key conceptual tools are:

- **Newton–Raphson root-finding** locates equilibria numerically
- **The Jacobian and its eigenvalues** determine local stability
- **The trace-determinant plane** organizes all 2D equilibrium types
- **Stable/unstable manifolds** of saddles organize the global phase
  portrait into basins of attraction

Together these tools provide a complete qualitative picture of an
autonomous ODE system's behavior without requiring closed-form
solutions.

## References

Gardner TS, Cantor CR, Collins JJ (2000). Construction of a genetic
toggle switch in *Escherichia coli*. *Nature* **403**: 339–342.

Grayling MJ (2014). phaseR: An R Package for Phase Plane Analysis of
Autonomous ODE Systems. *The R Journal* **6**(2): 43–51.

Strogatz SH (2015). *Nonlinear Dynamics and Chaos* (2nd ed.).
Westview Press.
