---
title: "Introduction to ggphasr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to ggphasr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width  = 6,
  fig.height = 5,
  fig.align  = "center",
  out.width  = "85%"
)
```

## What is ggphasr?

`ggphasr` provides tools for **qualitative analysis of one- and
two-dimensional autonomous ordinary differential equation (ODE) systems**
using phase plane methods. It is a modern successor to the
[phaseR](https://github.com/mjg211/phaseR) package (Grayling, 2014),
with two key improvements:

1. All visualizations are produced with **ggplot2** and its extensions,
   making every plot fully customizable with standard ggplot2 syntax.
2. The package is distributed via **GitHub** rather than CRAN, ensuring
   it remains installable regardless of CRAN archival policies.

`ggphasr` is designed for **students and instructors** in applied
mathematics, mathematical modeling, differential equations, and
dynamical systems courses. Plots compose naturally using the `+`
operator — the same way all ggplot2 plots are built — so students
already familiar with ggplot2 can immediately adapt and extend any
phase portrait.

```{r load, message = FALSE}
library(ggphasr)
library(ggplot2)
```

---

## Installation

```r
# install.packages("remotes")
remotes::install_github("jmgraham30/ggphasr")
```

---

## ODE conventions

`ggphasr` accepts ODE functions in two calling conventions.

**Convention A** — deSolve-compatible (recommended):

```r
my_ode <- function(t, y, parameters) {
  # y[1] = x, y[2] = y (for 2D systems)
  # parameters = named numeric vector
  list(c(
    f(y[1], y[2]),   # dx/dt
    g(y[1], y[2])    # dy/dt
  ))
}
```

**Convention B** — simplified, for quick exploration:

```r
# 2D:
my_ode <- function(x, y, parameters = NULL) c(dx, dy)

# 1D:
my_ode <- function(y, parameters = NULL) dy
```

The calling convention is detected automatically from the function's
argument names, so both styles work with all `ggphasr` functions.
Convention A is directly compatible with `deSolve::ode()`.

---

## The composable workflow

The core workflow builds a phase portrait layer by layer using `+`,
just like standard ggplot2:

```{r composable, message = FALSE}
lv_params <- c(alpha = 1, beta = 0.5, delta = 0.5, gamma = 1)

gg_flow_field(
  ode_lotka_volterra,
  xlim       = c(0, 5),
  ylim       = c(0, 5),
  parameters = lv_params,
  xlab       = "Prey",
  ylab       = "Predator",
  title      = "Lotka-Volterra predator-prey model"
) +
  gg_nullclines(
    ode_lotka_volterra,
    xlim            = c(0, 5),
    ylim            = c(0, 5),
    parameters      = lv_params,
    legend_position = "bottom"
  ) +
  gg_trajectory(
    ode_lotka_volterra,
    y0         = matrix(c(0.5, 0.5,
                          1.0, 3.0,
                          3.5, 0.5,
                          3.0, 3.5),
                        ncol = 2, byrow = TRUE),
    xlim       = c(0, 5),
    ylim       = c(0, 5),
    parameters = lv_params,
    t_end      = 25,
    color      = "grey30",
    add_start_point = FALSE
  )
```

Each function in this chain has a specific role:

- `gg_flow_field()` — creates the base ggplot object with direction
  arrows showing the vector field $(dx/dt, dy/dt)$ at each grid point.
  It returns a complete `ggplot` object.
- `gg_nullclines()` — adds the x-nullcline (where $dx/dt = 0$, red)
  and y-nullcline (where $dy/dt = 0$, blue). Returns a list of ggplot2
  layers.
- `gg_trajectory()` — numerically integrates the ODE from the supplied
  initial conditions using `deSolve` and adds the solution paths.
  Returns a list of ggplot2 layers.

Because `gg_nullclines()` and `gg_trajectory()` return layer lists
rather than complete ggplot objects, they compose naturally with `+`
just as any ggplot2 geom would.

---

## The all-in-one wrapper

For quick exploration, `gg_phase_plane()` produces a complete phase
portrait — flow field, nullclines, trajectories from a grid of initial
conditions, and classified equilibria — in a single call:

```{r phase-plane, message = FALSE}
result <- gg_phase_plane(
  ode_lotka_volterra,
  xlim            = c(0, 5),
  ylim            = c(0, 5),
  parameters      = lv_params,
  legend_position = "bottom"
)

result$plot
```

`gg_phase_plane()` returns a named list with two components:

- `$plot` — the ggplot object, which can be further customized with `+`
- `$equilibria` — a tidy data frame of classified equilibria

```{r equilibria}
result$equilibria[, c("x", "y", "classification", "tr", "det")]
```

The `add_layer()` function adds ggplot2 layers to the result while
preserving `$equilibria`:

```{r add-layer}
add_layer(result, labs(title = "Lotka-Volterra: classified equilibria"))$plot
```

---

## 1D systems

For one-dimensional systems, `system = "one.dim"` switches all
functions to phase line mode. `gg_phase_portrait()` adds the
traditional phase line with directional arrows and stability markers:

```{r one-dim, message = FALSE}
gg_flow_field(
  ode_logistic,
  xlim       = c(0, 6),
  ylim       = c(-1, 12),
  system     = "one.dim",
  parameters = c(r = 1, K = 10),
  xlab       = "Time t",
  ylab       = "Population y",
  title      = "Logistic growth: 1D phase portrait"
) +
  gg_nullclines(
    ode_logistic,
    xlim       = c(0, 6),
    ylim       = c(-1, 12),
    system     = "one.dim",
    parameters = c(r = 1, K = 10)
  ) +
  gg_trajectory(
    ode_logistic,
    y0         = list(c(0.5), c(3), c(7), c(11.5)),
    xlim       = c(0, 6),
    ylim       = c(-1, 12),
    system     = "one.dim",
    parameters = c(r = 1, K = 10),
    t_end      = 5,
    color      = "steelblue"
  ) +
  gg_phase_portrait(
    ode_logistic,
    ylim       = c(-1, 12),
    xlim       = c(0, 6),
    parameters = c(r = 1, K = 10)
  )
```

---

## Equilibrium analysis

`find_equilibrium()` and `classify_equilibrium()` provide numerical
tools for locating and classifying equilibria:

```{r equilibrium-analysis, message = FALSE}
# Find equilibria by grid search
eq_list <- find_equilibrium(
  ode_lotka_volterra,
  y0         = NULL,
  xlim       = c(0, 5),
  ylim       = c(0, 5),
  parameters = lv_params
)

# Classify each equilibrium
do.call(rbind, lapply(eq_list, function(eq) {
  classify_equilibrium(ode_lotka_volterra,
                       equilibrium = eq,
                       parameters  = lv_params)
}))[, c("x", "y", "classification", "tr", "det")]
```

---

## Stable and unstable manifolds

`gg_manifolds()` draws the stable and unstable manifolds of saddle
points, which organize the phase plane into basins of attraction:

```{r manifolds, message = FALSE}
eq    <- find_equilibrium(ode_example_11, y0 = c(0.8, 0.8))
eq_cl <- classify_equilibrium(ode_example_11, equilibrium = eq[[1L]])

gg_flow_field(ode_example_11,
              xlim  = c(0, 4),
              ylim  = c(0, 3),
              title = "Competition model: saddle manifolds") +
  gg_nullclines(ode_example_11,
                xlim            = c(0, 4),
                ylim            = c(0, 3),
                legend_position = "bottom") +
  gg_manifolds(ode_example_11,
               equilibrium   = eq[[1L]],
               eq_classified = eq_cl,
               t_manifold    = 5)
```

---

## Built-in ODE systems

`ggphasr` includes `r 4 + 8 + 15` built-in ODE systems ready to use
without writing any equations:

**1D growth models** (4):
`ode_exponential()`, `ode_logistic()`, `ode_monomolecular()`,
`ode_von_bertalanffy()`

**Classical 2D models** (8):
`ode_lotka_volterra()`, `ode_sir()`, `ode_van_der_pol()`,
`ode_simple_pendulum()`, `ode_competition()`, `ode_toggle()`,
`ode_morris_lecar()`, `ode_lindemann()`

**Textbook examples** (15):
`ode_example_01()` through `ode_example_15()` — a library of 1D and
2D systems covering all common equilibrium types (stable/unstable
nodes, spirals, centers, saddles, and limit cycles).

All built-in systems use Convention A and can be passed directly to
any `ggphasr` plotting or analysis function.

---

## Function reference

### Plotting

| Function | Returns | Description |
|---|---|---|
| `gg_flow_field()` | `ggplot` | Direction/velocity field |
| `gg_nullclines()` | layer list | Nullcline curves |
| `gg_trajectory()` | layer list | Solution trajectories |
| `gg_phase_portrait()` | layer list | 1D phase line |
| `gg_time_series()` | `ggplot` | Time-series plots |
| `gg_manifolds()` | layer list | Stable/unstable manifolds |
| `gg_phase_plane()` | `ggphasr_result` | All-in-one wrapper |

### Analysis

| Function | Returns | Description |
|---|---|---|
| `find_equilibrium()` | list | Newton–Raphson root-finding |
| `classify_equilibrium()` | data frame | Stability classification |
| `add_layer()` | `ggphasr_result` | Add layers to a result object |

### Theme

| Function | Returns | Description |
|---|---|---|
| `theme_phase_plane()` | `theme` | ggplot2 theme for phase portraits |

---

## Further reading

The three topic vignettes provide full mathematical exposition and
detailed worked examples:

```r
# One-dimensional systems: phase lines, logistic growth, stability
vignette("one-dimensional-systems", package = "ggphasr")

# Two-dimensional systems: phase planes, nullclines, limit cycles
vignette("two-dimensional-systems", package = "ggphasr")

# Equilibrium analysis: Jacobian, trace-determinant, manifolds
vignette("equilibrium-analysis", package = "ggphasr")
```

---

## References

Grayling MJ (2014). phaseR: An R Package for Phase Plane Analysis of
Autonomous ODE Systems. *The R Journal* **6**(2): 43–51.
<https://doi.org/10.32614/RJ-2014-023>

Strogatz SH (2015). *Nonlinear Dynamics and Chaos* (2nd ed.).
Westview Press.
