---
title: "One-Dimensional ODE Systems"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{One-Dimensional ODE Systems}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 6,
  fig.height = 4,
  fig.align  = "center",
  out.width  = "90%"
)
```

## Introduction

A **one-dimensional autonomous ODE** has the form

$$\frac{dy}{dt} = f(y),$$

where the derivative depends only on the current state $y$, not on time
$t$ directly.  Despite their simplicity, 1D autonomous systems already
exhibit rich qualitative behavior: exponential growth or decay, logistic
saturation, bistability, and more.

**Phase line analysis** is the standard tool for understanding these
systems without solving them explicitly.  The key idea is to plot $f(y)$
as a function of $y$, identify where $f(y) = 0$ (the equilibria), and
read off the direction of flow from the sign of $f(y)$.

This vignette shows how to carry out a complete 1D phase analysis using
`ggphasr`.  We cover:

1. Defining 1D ODE systems in `ggphasr`
2. Plotting the flow field and phase line
3. Adding nullclines and trajectories
4. Using `find_equilibrium()` and `classify_equilibrium()`
5. The `gg_phase_plane()` all-in-one wrapper
6. Worked examples: logistic growth, a user-defined cubic, and the
   von Bertalanffy growth model

```{r load-packages, message = FALSE}
library(ggphasr)
library(ggplot2)
```

---

## 1  Defining a 1D ODE in ggphasr

`ggphasr` accepts ODE functions in two conventions.

**Convention A** (deSolve-compatible, recommended):

```r
f <- function(t, y, parameters) {
  r <- parameters[["r"]]
  K <- parameters[["K"]]
  list(c(r * y[1] * (1 - y[1] / K)))
}
```

- First argument is time `t` (present for deSolve compatibility, but
  ignored in autonomous systems).
- Second argument `y` is a numeric vector of state variables (length 1
  for 1D systems).
- Third argument `parameters` is a named numeric vector of model
  parameters.
- Return value is a **list** whose first element is a numeric vector of
  derivatives.

**Convention B** (simplified, for quick exploration):

```r
f <- function(y, parameters = NULL) {
  r * y * (1 - y / K)
}
```

- No time argument.
- Returns a plain numeric scalar, not a list.
- `ggphasr` detects Convention B automatically and wraps it into
  Convention A internally.

> **Note for advanced users:** Convention A is directly compatible with
> `deSolve::ode()`, which underpins all numerical integration in
> `ggphasr`.  Convention B is a convenience wrapper; under the hood it
> is converted to Convention A before any computation.

Throughout this vignette we use the built-in `ode_logistic()` and
define a few custom systems to illustrate both conventions.

---

## 2  The logistic growth model

The logistic (Verhulst) model is the classic 1D ODE for bounded
population growth:

$$\frac{dy}{dt} = r\,y\!\left(1 - \frac{y}{K}\right),$$

where $y(t)$ is population size, $r > 0$ is the intrinsic growth rate,
and $K > 0$ is the carrying capacity.

The model has two equilibria: $y^* = 0$ (unstable — populations
near zero grow away from it) and $y^* = K$ (stable — populations near
$K$ return to it).  The analytic solution is the familiar S-shaped
logistic curve

$$y(t) = \frac{K}{1 + \left(\frac{K - y_0}{y_0}\right)e^{-rt}}.$$

### 2.1  Flow field

`gg_flow_field()` with `system = "one.dim"` plots the direction field:
arrows in the $(t, y)$ plane showing whether $y$ is increasing or
decreasing at each point.

```{r logistic-flow-field}
gg_flow_field(
  ode_logistic,
  xlim       = c(0, 8),
  ylim       = c(-1, 12),
  system     = "one.dim",
  parameters = c(r = 1, K = 10),
  xlab       = "Time t",
  ylab       = "Population y",
  title      = "Logistic growth: flow field"
)
```

Notice that arrows point upward (y increasing) for $0 < y < 10$ and
downward (y decreasing) for $y > 10$, reflecting the two equilibria.

### 2.2  Adding the phase line

`gg_phase_portrait()` adds the traditional phase line — a vertical
reference line with directional arrows and stability markers:

- **Filled circle** = stable equilibrium (arrows point toward it)
- **Open circle** = unstable equilibrium (arrows point away from it)

```{r logistic-phase-portrait}
gg_flow_field(
  ode_logistic,
  xlim       = c(0, 8),
  ylim       = c(-1, 12),
  system     = "one.dim",
  parameters = c(r = 1, K = 10),
  xlab       = "Time t",
  ylab       = "Population y",
  title      = "Logistic growth: flow field + phase line"
) +
  gg_phase_portrait(
    ode_logistic,
    ylim       = c(-1, 12),
    xlim       = c(0, 8),
    parameters = c(r = 1, K = 10)
  )
```

### 2.3  Adding nullclines and trajectories

The nullcline of a 1D system is simply the set of $y$ values where
$f(y) = 0$ — the equilibrium points.  `gg_nullclines()` draws these as
horizontal lines.  `gg_trajectory()` integrates the ODE forward in time
from specified initial conditions.

```{r logistic-full}
gg_flow_field(
  ode_logistic,
  xlim       = c(0, 8),
  ylim       = c(-1, 12),
  system     = "one.dim",
  parameters = c(r = 1, K = 10),
  xlab       = "Time t",
  ylab       = "Population y",
  title      = "Logistic growth: complete phase portrait"
) +
  gg_nullclines(
    ode_logistic,
    xlim            = c(0, 8),
    ylim            = c(-1, 12),
    system          = "one.dim",
    parameters      = c(r = 1, K = 10),
    legend_position = "bottom"
  ) +
  gg_trajectory(
    ode_logistic,
    y0         = list(c(0.5), c(3), c(7), c(11.5)),
    xlim       = c(0, 8),
    ylim       = c(-1, 12),
    system     = "one.dim",
    parameters = c(r = 1, K = 10),
    t_end      = 7,
    color      = "steelblue"
  ) +
  gg_phase_portrait(
    ode_logistic,
    ylim       = c(-1, 12),
    xlim       = c(0, 8),
    parameters = c(r = 1, K = 10)
  )
```

Four trajectories are shown: two starting below $K$ (growing toward it),
one starting above $K$ (declining toward it), and one starting near zero.
All converge to the stable equilibrium $y^* = K = 10$.

### 2.4  Time series

`gg_time_series()` shows the same solutions plotted as $y$ vs $t$,
which is the more familiar representation in many courses:

```{r logistic-time-series}
gg_time_series(
  ode_logistic,
  y0         = list(c(0.5), c(3), c(7), c(11.5)),
  t_end      = 8,
  system     = "one.dim",
  parameters = c(r = 1, K = 10),
  xlab       = "Time t",
  ylab       = "Population y(t)",
  title      = "Logistic growth: time series"
)
```

---

## 3  Equilibrium analysis

### 3.1  Finding equilibria numerically

`find_equilibrium()` locates equilibria using Newton–Raphson
root-finding.  For the logistic equation we know the equilibria are
$y^* = 0$ and $y^* = K = 10$, so we can verify:

```{r logistic-find-eq}
eq_list <- find_equilibrium(
  ode_logistic,
  y0         = NULL,       # grid search over ylim
  system     = "one.dim",
  ylim       = c(-0.5, 11),
  parameters = c(r = 1, K = 10)
)
eq_list
```

### 3.2  Classifying equilibria

`classify_equilibrium()` computes $f'(y^*)$ via finite differences and
uses its sign to determine stability:

$$\frac{df}{dy}\bigg|_{y^*} < 0 \implies \text{stable}, \qquad
\frac{df}{dy}\bigg|_{y^*} > 0 \implies \text{unstable}.$$

For the logistic equation, $f(y) = ry(1 - y/K)$, so
$f'(y) = r(1 - 2y/K)$.  At $y^* = 0$: $f'(0) = r > 0$ (unstable).
At $y^* = K$: $f'(K) = -r < 0$ (stable).

```{r logistic-classify}
eq_table <- do.call(rbind, lapply(eq_list, function(eq) {
  classify_equilibrium(ode_logistic,
                       equilibrium = eq,
                       system      = "one.dim",
                       parameters  = c(r = 1, K = 10))
}))
eq_table[, c("y", "classification", "jacobian_11")]
```

---

## 4  User-defined example: a cubic ODE

To illustrate working with a user-defined system, consider the cubic:

$$\frac{dy}{dt} = y(1 - y)(2 - y),$$

which appears as `ode_example_02()` in `ggphasr`.  This system has
three equilibria at $y^* = 0$, $1$, and $2$ with alternating stability
(unstable, stable, unstable as $y$ increases).

We can define this in Convention B for conciseness:

```{r cubic-define}
cubic_ode <- function(y, parameters = NULL) {
  y * (1 - y) * (2 - y)
}
```

> **Convention B in action:** `ggphasr` detects that the first argument
> is named `y` (not `t`) and automatically wraps this function into
> Convention A before passing it to deSolve.

```{r cubic-portrait}
gg_flow_field(
  cubic_ode,
  xlim   = c(0, 5),
  ylim   = c(-0.5, 2.5),
  system = "one.dim",
  xlab   = "Time t",
  ylab   = "y",
  title  = "Cubic ODE: dy/dt = y(1-y)(2-y)"
) +
  gg_nullclines(
    cubic_ode,
    xlim   = c(0, 5),
    ylim   = c(-0.5, 2.5),
    system = "one.dim"
  ) +
  gg_trajectory(
    cubic_ode,
    y0     = list(c(-0.3), c(0.4), c(1.5), c(2.3)),
    xlim   = c(0, 5),
    ylim   = c(-0.5, 2.5),
    system = "one.dim",
    t_end  = 4,
    color  = "grey30"
  ) +
  gg_phase_portrait(
    cubic_ode,
    ylim = c(-0.5, 2.5),
    xlim = c(0, 5)
  )
```

```{r cubic-classify}
eq_cubic <- find_equilibrium(cubic_ode,
                              y0     = NULL,
                              system = "one.dim",
                              ylim   = c(-0.5, 2.5))

do.call(rbind, lapply(eq_cubic, function(eq) {
  classify_equilibrium(cubic_ode, equilibrium = eq, system = "one.dim")
}))[, c("y", "classification", "jacobian_11")]
```

As expected: $y^* = 0$ is unstable ($f'(0) > 0$), $y^* = 1$ is stable
($f'(1) < 0$), and $y^* = 2$ is unstable ($f'(2) > 0$).

---

## 5  The all-in-one wrapper: gg_phase_plane()

For quick exploration, `gg_phase_plane()` produces the complete phase
portrait — flow field, nullclines, trajectories, and classified
equilibria — in a single call.  It returns a named list with `$plot`
(the ggplot object) and `$equilibria` (the classification table).

```{r cubic-phase-plane}
result <- gg_phase_plane(
  cubic_ode,
  xlim            = c(0, 5),
  ylim            = c(-0.5, 2.5),
  system          = "one.dim",
  t_end           = 4,
  title           = "Cubic ODE: gg_phase_plane() output",
  legend_position = "bottom"
)

result$plot
result$equilibria[, c("y", "classification")]
```

---

## 6  Von Bertalanffy growth model

The von Bertalanffy model describes organism growth in terms of
anabolism (proportional to surface area $\sim y^{2/3}$) and catabolism
(proportional to mass $\sim y$):

$$\frac{dy}{dt} = \alpha\, y^{2/3} - \beta\, y,$$

where $y(t)$ is body mass, $\alpha > 0$ is the anabolism coefficient,
and $\beta > 0$ is the catabolism coefficient.  Setting $dy/dt = 0$
gives the stable asymptotic mass

$$y^* = \left(\frac{\alpha}{\beta}\right)^3.$$

With $\alpha = 1$ and $\beta = 0.5$, we get $y^* = (1/0.5)^3 = 8$.

> **Technical aside:** The $y^{2/3}$ term means $f(y)$ is not
> differentiable at $y = 0$, so the trivial equilibrium at $y = 0$
> requires care.  `classify_equilibrium()` will return
> `"Inconclusive (df/dy = 0)"` there because the finite-difference
> Jacobian is undefined.  This is expected and biologically
> unimportant — organisms start with positive mass.

```{r vb-portrait}
gg_flow_field(
  ode_von_bertalanffy,
  xlim       = c(0, 6),
  ylim       = c(0, 14),
  system     = "one.dim",
  parameters = c(alpha = 1, beta = 0.5),
  xlab       = "Time t",
  ylab       = "Body mass y",
  title      = "Von Bertalanffy growth (alpha=1, beta=0.5)"
) +
  gg_nullclines(
    ode_von_bertalanffy,
    xlim       = c(0, 6),
    ylim       = c(0, 14),
    system     = "one.dim",
    parameters = c(alpha = 1, beta = 0.5)
  ) +
  gg_trajectory(
    ode_von_bertalanffy,
    y0         = list(c(1), c(4), c(8), c(12)),
    xlim       = c(0, 6),
    ylim       = c(0, 14),
    system     = "one.dim",
    parameters = c(alpha = 1, beta = 0.5),
    t_end      = 5,
    color      = "steelblue"
  ) +
  gg_phase_portrait(
    ode_von_bertalanffy,
    ylim       = c(0, 14),
    xlim       = c(0, 6),
    parameters = c(alpha = 1, beta = 0.5)
  ) +
  # Mark the asymptotic mass y* = 8 analytically
  geom_hline(yintercept = 8, linetype = "dotted", color = "firebrick") +
  annotate("text", x = 5.5, y = 8.4,
           label = "y* = 8", color = "firebrick", size = 3.5)
```

The dotted horizontal line marks the analytic equilibrium $y^* = 8$.
All trajectories — regardless of starting mass — converge to it,
confirming that $y^*$ is globally stable in the region $y > 0$.

---

## 7  Comparing growth models

A strength of `ggphasr`'s composable design is that comparing systems
side by side requires only swapping the ODE function.  Here we compare
the time series of three growth models with the same carrying capacity
$K = 10$:

```{r compare-growth}
t_end  <- 10
params <- c(r = 1, K = 10)
y0     <- list(c(1))

p_exp  <- gg_time_series(ode_exponential,   y0 = y0, t_end = t_end,
                          system = "one.dim", parameters = c(r = 0.5),
                          title = "Exponential")$plot +
  coord_cartesian(ylim = c(0, 15))

p_log  <- gg_time_series(ode_logistic,       y0 = y0, t_end = t_end,
                          system = "one.dim", parameters = params,
                          title = "Logistic")$plot

p_mono <- gg_time_series(ode_monomolecular,  y0 = y0, t_end = t_end,
                          system = "one.dim", parameters = params,
                          title = "Monomolecular")$plot
```

```{r compare-growth-plot, fig.width = 9, fig.height = 3.5}
# Combine with patchwork if available, otherwise view separately
if (requireNamespace("patchwork", quietly = TRUE)) {
  library(patchwork)
  p_exp | p_log | p_mono
} else {
  p_log
}
```

> **Key differences:** Exponential growth is unbounded (no carrying
> capacity).  Logistic growth has an S-shaped curve with an inflection
> point at $y = K/2$.  Monomolecular growth is concave throughout —
> the growth rate decreases monotonically without an inflection point.

---

## Summary

This vignette covered the complete `ggphasr` workflow for 1D autonomous
ODE systems:

| Task | Function |
|---|---|
| Plot direction field | `gg_flow_field(..., system = "one.dim")` |
| Add nullclines | `gg_nullclines(..., system = "one.dim")` |
| Add trajectories | `gg_trajectory(..., system = "one.dim")` |
| Add phase line | `gg_phase_portrait()` |
| Plot time series | `gg_time_series(..., system = "one.dim")` |
| Find equilibria | `find_equilibrium(..., system = "one.dim")` |
| Classify equilibria | `classify_equilibrium(..., system = "one.dim")` |
| All-in-one | `gg_phase_plane(..., system = "one.dim")` |

The next vignette, *Two-Dimensional ODE Systems*, extends these ideas
to 2D systems where trajectories move in the $(x, y)$ phase plane and
equilibria are classified by the full Jacobian matrix.
